Reconstructing Propositional Proofs in Type Theory

### 2017-11-17 paper/sections/type-theory.tex
[ paper ] some corrections added.

Type theory is a formalism for the foundation of mathematics, and it has became in a key instrument to study logic and proof theory.
It was proposed in 1902 to avoid Russell's paradox and it follows
the same philosophy of constructive mathematics where the witness of
a statement is \emph{everything}. In that direction, we could say
the main actors in type theory are the statements and their proofs.
By following the Curry-Howard correspondence (see, for example,~\cite{Wadler-2015}), we say
\emph{types} are theorems and their \emph{terms} are their proofs.

@@

We produce derivation trees inside the judgments using the derivation rules.
Since type theory can be seen as typed $λ$-calculus with dependent function
types, evaluation of $λ$-terms also called \emph{normalization} of the proof-
terms, are reductions using the system inference rules. Therefore, the task to
verify a proof that proves a certain theorem is translated to verify if the
associated tree of the proof has the correspondent type of the theorem. We know
this process as \emph{type-checking}. This feature of type theory allow us to
verify a proof generated by an \ATP by reconstructing its proof-term to type-
check the proof. For such a purpose, we use a proof assistant based on type
theory like \Agda to delegate this task.
% However, this approach has some weak points discussed in~\cite{Bezem2002}.
Some limitations from the type theory
point of view for proof-reconstruction is described in the following sections.

@@

Functions defined by a general recursion can not be directly translated in type
theory since it is not a guarantee they terminate. For that reason, we follow
the technique described in~\cite{Bertot2004} to avoid termination problems by
modifying the recursive functions to be \emph{structural recursive}. % (see
\cite{Coquand1992,Abel2002,Bove2005} for more details).

@@

\begin{notation}
  We use \Prop for the type of propositions.  For syntactical
  equivalence, we write the equality as $φ ≡ ψ$, for $φ,~ψ :
  \Prop$. For this equality, we assume the reflexivity, symmetry and
  transitivity properties. The type of natural numbers is called \Nat,
  and it is defined as usual, where $\zero$ and $\suc$ are its data
  constructors. We use names and symbols for the arithmetic operations
  as usual (\eg, $+$, $-$, $*$).  We use syntax sugar for $\zero$, $\suc$,
  $\suc~\zero$, $\dots$, with the decimal representation~$0$, $1$,
  $2$, $\dots$ as well.
\end{notation}

@@

In~\eqref{eq:uh-definition}, the first two equations
call on arguments no structurally smaller.
For instance, on the first call, $φ ≡ (φ₁ ∧ φ₂) ⇒ φ₃$, the recursion
calls on the  $φ₁ ⇒ (φ₂ ⇒ φ₃)$ argument that is not a subformula of φ.
Therefore, one way to translate $\fuh_{0}$ to type
theory is to define a new function using a bounded recursion.

@@

Type theory and intuitionistic logic are accompanying theories since
they are based on the same basis of constructive mathematics. Proving
propositions in these theories demands a witness construction for the
proof. Nonetheless, classical logic does not always have constructive
proofs and some of its proofs are proved by non-constructive methods
like refutation.

To reconstruct proofs generated by \Metis, we have formalized in type
theory the classical propositional logic in~\cite{AgdaProp}.  A
briefly description of it is presented in \cite{VanDalen1994}. In this
formalization, we have to assume the principle of excluded third
(henceforth \abbre{PEM}) as an axiom since \Metis is a prover for
classical logic. Assuming \abbre{PEM}, we can justify refutation
proofs by deriving from it the \emph{reductio ad absurdum} rule
(henceforth \abbre{RAA}). The \abbrev{RAA} rule is the formulation of
the principle of proof by contradiction, that is, a derivation of a
contradiction,~$⊥$, from the hypothesis~$¬ φ$, is a derivation of~$φ$.

>>>>

Type theory is a formalism for the foundation of mathematics, and it has became in a key instrument to study logic and proof theory.
It was proposed in 1902 to avoid Russell's paradox and it follows
the same philosophy of constructive mathematics where the witness of
a statement is \emph{everything}. In that direction, we could say
the main actors in type theory are the statements and their proofs.
By following the Curry-Howard correspondence (see, for example,~\cite{Wadler-2015}), types as propositions.
A proposition corresponds to a type, and a proof of that propositions
corresponds to the term of that proof (henceforth proof-term).
%\emph{types} are theorems and their \emph{terms} are their proofs.

@@

We produce derivation trees inside the judgments using the derivation rules.
Since type theory can be seen as typed $λ$-calculus with dependent function
types, evaluation of $λ$-terms also called \emph{normalization} of the proof-
terms, are reductions using the system inference rules.

Therefore, the
proof verification task becomes in type theory as verifying that the proof-term has the correspondent type of the theorem. We know
this process as \emph{type-checking}. This feature of type theory allow us to
verify a proof generated by an \ATP by reconstructing its proof-term to type-
check the proof. For such a purpose, we use a proof assistant based on type
theory like \Agda to delegate this task.
% However, this approach has some weak points discussed in~\cite{Bezem2002}.
Some limitations from the type theory
point of view for proof-reconstruction is described in the following sections.

@@

Functions defined by a general recursion can not be directly translated in type
theory since it is not a guarantee they terminate. For that reason, we follow
the technique described in~\cite{Bertot2004} to avoid termination problems by
modifying the recursive functions to be \emph{structurally recursive}. % (see
\cite{Coquand1992,Abel2002,Bove2005} for more details).

@@

\begin{notation}
  We use \Prop for the type of propositions.
  We write the equality as $φ ≡ ψ$, for $φ,~ψ :
  \Prop$. For this equality, we assume the reflexivity, symmetry and
  transitivity properties.
  We say $\varphi$ and $\psi$ are Two propositions are syntactically equivalent if there is a derivation from $\varphi$ to $\psi$ and
  a derivation from $\psi$ to $\varphi$.
   The type of natural numbers is called \Nat,
  and it is defined as usual, where $\zero$ and $\suc$ are its data
  constructors. We use names and symbols for the arithmetic operations
  as usual (\eg, $+$, $-$, $*$).  We use syntax sugar for $\zero$, $\suc$,
  $\suc~\zero$, $\dots$, with the decimal representation~$0$, $1$,
  $2$, $\dots$ as well.
\end{notation}

@@

In~\eqref{eq:uh-definition}, the first two equations
call on arguments no structurally smaller.
For instance, in the first equation,
$$\fuh_{0} (φ₁ ⇒ (φ₂ ⇒ φ₃)) = \fuh_{0}~((φ₁ ∧ φ₂) ⇒ φ₃),$$
the function $\fuh_{0}$ calls on the  $(φ₁ ∧ φ₂) ⇒ φ₃$ argument in the right-hand side but this formula is not a subformula of $φ₁ ⇒ (φ₂ ⇒ φ₃)$.
Therefore, one way to translate $\fuh_{0}$ to type
theory is to define a new function using a bounded recursion.

@@

Type theory and intuitionistic logic are accompanying theories since
they are based on the same basis of constructive mathematics. Proving
propositions in these theories demands a witness construction for the
proof. Nonetheless, classical logic does not always have constructive
proofs since some of the proofs are stated by refutation.

To reconstruct proofs generated by \Metis, we have formalized in type
theory the classical propositional logic in~\cite{AgdaProp}.  A
briefly description of it is presented in \cite{VanDalen1994}. In this
formalization, we have to assume the principle of excluded middle
(henceforth \abbre{PEM}) as an axiom since \Metis is a prover for
classical logic. Assuming \abbre{PEM}, we can justify refutation
proofs by deriving from it the \emph{reductio ad absurdum} rule
(henceforth \abbre{RAA}). The \abbrev{RAA} rule is the formulation of
the principle of proof by contradiction, that is, a derivation of a
contradiction,~$⊥$, from the hypothesis~$¬ φ$, is a derivation of~$φ$.
